---
title: "Bayesian Online Changepoint Detection in bocpdGraph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Online Changepoint Detection in bocpdGraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(bocpdGraph)
```

## Overview

The **bocpdGraph** package implements Bayesian Online Changepoint Detection
(BOCPD) following Adams and MacKay (2007). The implementation adheres to the
original two-branch run-length recursion while introducing safeguards required
for numerical stability, extensibility, and long-running use.

The package is organized around four core components:

- Hazard functions (`hazard_*`)
- Run-length recursion (internal, log scale)
- The BOCPD update engine (`bocpd_engine`)
- A user-facing interface (`bocpd`)

Auxiliary helpers enforce probabilistic invariants and control computational
complexity.

---

## Run-length recursion (internal)

At time \(t\), the run length \(r_t\) denotes the number of observations since
the most recent changepoint. Internally, **bocpdGraph** represents the run-length
posterior as a numeric vector on the log scale,

\[
\log P(r_t \mid y_{1:t}), \quad r_t = 0, 1, \dots
\]

rather than as a user-facing object.

Run-length updates are performed entirely in log space. The recursion implements:

- **growth:** \(r \to r+1\) with probability \(1 - h(r, t)\)
- **changepoint:** \(r \to 0\) with probability \(h(r, t)\)

Likelihood evaluation is supplied externally by the observation model; the
run-length utilities are likelihood-agnostic and graph-agnostic.

Optional truncation and pruning approximate the exact recursion while preserving
its probabilistic interpretation.

---

## Hazard functions

### Concept

The hazard function specifies the prior probability of a changepoint conditional
on the current run length:

\[
H(r, t) = P(r_t = 0 \mid r_{t-1} = r).
\]

In **bocpdGraph**, hazards are supplied as functions of the form

```
h(run_length, t, ...)
```

returning probabilities in \([0,1]\). Several common hazard families are
provided.

### Constant (geometric) hazard

```{r}
h_const <- hazard_constant(0.01)
h_const(0:5, t = 10)
```

### Time-varying hazard

```{r}
h_time <- hazard_piecewise_time(
  times = c(50, 100),
  p = c(0.01, 0.05, 0.02)
)

h_time(0:3, t = 25)
h_time(0:3, t = 80)
```

### Run-length–dependent hazard

```{r}
h_rl <- hazard_logistic_run_length(a = -6, b = 0.05)
h_rl(0:10, t = 1)
```

---

## Observation model

### IID Gaussian mean model

For illustration, the package provides a minimal IID Gaussian mean model,
intended primarily for testing and examples.

```{r}
model <- gaussian_iid_model(mu0 = 0, sigma = 1)
```

This model assumes a known variance and updates only the posterior mean across
run lengths.

---

## BOCPD engine

### `bocpd_engine`

The function `bocpd_engine()` implements the BOCPD recursion independently of any
specific observation model. At each time step, the engine:

1. Evaluates predictive likelihoods
2. Updates the run-length posterior
3. Normalizes probabilities
4. Advances model-specific sufficient statistics

The engine is designed for reuse in structured extensions such as graph-based or
multistream BOCPD.

```{r, eval=FALSE}
engine <- bocpd_engine(
  hazard = h_const,
  model = model
)
```

---

## User-facing interface

### `bocpd`

The function `bocpd()` provides a high-level interface for applying BOCPD to a
univariate time series.

```{r}
set.seed(1)
y <- c(rnorm(50), rnorm(50, mean = 3))

res <- bocpd(
  y,
  hazard = h_const,
  model = model
)
```

The returned object contains:

- Run-length posterior summaries
- Changepoint probabilities
- Model-specific posterior quantities

This interface is suitable for most applied use cases.

---

## Deviations from the idealized algorithm

The implementation follows Adams and MacKay (2007) with intentional deviations
motivated by numerical stability and software engineering considerations.

### Hazard interface

Hazards are evaluated as

```
h(run_length, t, ...)
```

rather than as functions of run length alone. This allows time-varying or
exogenous changepoint priors.

### Run-length truncation and pruning

Optional truncation (`r_max`) and probability pruning (`prune_eps`) are provided
to control computational complexity. These operations approximate the exact
recursion but do not alter the underlying probabilistic model.

### Numerical safeguards

When floating-point underflow or overflow prevents reliable normalization of the
run-length posterior, the implementation falls back to a valid distribution
placing all mass on \(r_t = 0\). This behavior preserves probabilistic
invariants but is not part of the mathematical BOCPD algorithm.

---

## References

Adams, R. P. and MacKay, D. J. C. (2007).
*Bayesian Online Changepoint Detection*. arXiv:0710.3742.

Bretz, F. (2021).
*Bayesian Online Changepoint Detection*. Lecture notes / technical report.

